---
title: "VDJdb paper figures"
author: "Mikhail Shugay"
date: "26-02-2022"
output:
  html_document:
    code_folding: hide
---

<!--
The figures of the 2022 VDJdb paper and the miscellaneous curation tables, split out of
`vdjdb_summary.Rmd` at its own `!summary_embed_end!` marker. They are NOT part of a release: nothing
here reaches `vdjdb_summary_embed.html`, and `vdjdb-web` never reads this file.

The split is what keeps `maps`, `scatterpie` and `ggrepel` out of the release render path -- they are
called only below, and `maps`/`scatterpie` are not even installed on the reference machine.
-->

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(knitr)
library(ggplot2)
library(RColorBrewer)
library(data.table)
library(ggrepel)
library(tidyverse)
library(stringr)
library(maps)
library(scatterpie)
select = dplyr::select

df = fread("../database/vdjdb.slim.txt", header=T, sep="\t")
```

---

#### VDJdb 2022 paper

Total COVID records

```{r}
df %>%
  filter(species == "HomoSapiens", 
         startsWith(as.character(antigen.species), "SARS-CoV")) %>%
  nrow()
```

Total COVID epitopes

```{r}
df %>%
  filter(species == "HomoSapiens", 
         startsWith(as.character(antigen.species), "SARS-CoV")) %>%
  .$antigen.epitope %>%
  unique %>% 
  length
```

According to VDJdb'19 "since the establishment of the database a total of 155 published studies were processed and added resulting in 61049 TCR specificity records in 2019 compared to only 5491 in 2017."

Count total new records

```{r}
nrow(df) - 61049
```

Total number of publications

```{r}
dt.pubdate$reference.id %>%
  unique %>%
  length
```

Number of epitopes ("Notably, while the number of unique epitopes with known specific TCRs increased, it is still dwarfed by the number of unique TCRs in the database, as there are currently only 212 epitopes in VDJdb. " for 2019)

```{r}
df %>%
  .$antigen.epitope %>%
  unique %>% 
  length
```

Numbers of records and epitopes for common viruses.

CMV

```{r}
df %>% filter(antigen.species == "CMV") %>% nrow
df %>% filter(antigen.species == "CMV") %>% .$antigen.epitope %>% unique %>% length
```

Flu

```{r}
df %>% filter(antigen.species == "InfluenzaA") %>% nrow
df %>% filter(antigen.species == "InfluenzaA") %>% .$antigen.epitope %>% unique %>% length
```

EBV

```{r}
df %>% filter(antigen.species == "EBV") %>% nrow
df %>% filter(antigen.species == "EBV") %>% .$antigen.epitope %>% unique %>% length
```

Count number of HLA haplotypes. Class I

```{r}
df %>%
  filter(species == "HomoSapiens", mhc.class == "MHCI") %>%
  mutate(allele = str_split_fixed(mhc.a, "[:,]", 2)[,1]) %>%
  .$allele %>%
  unique %>%
  length
```

class II

```{r}
df %>%
  filter(species == "HomoSapiens", mhc.class == "MHCII") %>%
  mutate(allele1 = str_split_fixed(mhc.a, "[:,]", 2)[,1],
         allele2 = str_split_fixed(mhc.b, "[:,]", 2)[,1],
         allele = paste(allele1, allele2)) %>%
  .$allele %>%
  unique %>%
  length
```

Count number of records for HLA-A*02

```{r}
df %>%
  mutate(allele = str_split_fixed(mhc.a, "[:,]", 2)[,1]) %>%
  filter(allele == "HLA-A*02") %>%
  nrow()
```

and HLA-A*03

```{r}
df %>%
  mutate(allele = str_split_fixed(mhc.a, "[:,]", 2)[,1]) %>%
  filter(allele == "HLA-A*03") %>%
  nrow()
```

Number of studies with self-antigens

```{r}
df %>%
  filter(antigen.species == "HomoSapiens") %>%
  .$reference.id %>%
  unique %>%
  length
```
number of records in them

```{r}
df %>%
  filter(antigen.species == "HomoSapiens") %>%
  nrow()
```

and the number of unique epitopes

```{r}
df %>%
  filter(antigen.species == "HomoSapiens") %>%
  .$antigen.epitope %>%
  unique %>%
  length
```

COVID19 - number of records in original Shomuradova paper

```{r}
df %>%
  filter(reference.id == "https://github.com/antigenomics/vdjdb-db/issues/315") %>%
  nrow
```

total records by 2022

```{r}
df %>%
  filter(antigen.species == "SARS-CoV-2") %>%
  nrow
```

and number of epitopes

```{r}
df %>%
  filter(antigen.species == "SARS-CoV-2") %>%
  .$antigen.epitope %>%
  unique %>%
  length
```

Figure1b and TableS1

```{r}
pdf("covid_alluvium.pdf", width = 12, height = 10)
p1
dev.off()

df.c %>% reshape2::dcast(Gene + HLA + Epitope + Studies ~ `TCR chain`, fill = 0) %>%
        arrange(-(TRB+TRA)) %>%
  fwrite("covid_summary.txt", sep = "\t")
```

```{r fig.width=9, fig.height=4.5}
df %>%
  filter(antigen.species == "SARS-CoV-2") %>%
  select(antigen.epitope, reference.id) %>%
  separate_rows(reference.id, sep=",") %>%
  group_by(reference.id, antigen.epitope) %>%
  summarise(count = n()) %>%
  group_by(reference.id) %>%
  mutate(total = sum(count)) -> cov19.by.lab

#unique(cov19.by.lab$reference.id)

cov19.by.lab.coords <- tibble(
  reference.id = c(
    "https://doi.org/10.1101/2020.05.04.20085779",        
    "https://github.com/antigenomics/vdjdb-db/issues/315",
    "https://github.com/antigenomics/vdjdb-db/issues/323",
    "https://github.com/antigenomics/vdjdb-db/issues/326",
    "https://github.com/antigenomics/vdjdb-db/issues/333",
    "PMID:33664060",                                      
    "PMID:33951417",                                      
    "PMID:34061349",                                      
    "PMID:34086357",                                      
    "PMID:34222571"),
  city = c("Seattle",
           "Moscow",
           "Cardiff",
           "Memphis",
           "Moscow",
           "Vienna",
           "Melbourne",
           "Erlangen",
           "Melbourne",
           "Chongqing"),
  long = c(
    -122.3351,        
    37.6184,
    -3.1790,
    -89.9711,
    37.6184,
    16.3634, 
    144.9464,                                      
    11.0166,                                      
    144.9464,                                      
    106.8877),
  lat = c(
    47.6080,        
    55.7512,
    51.4815,
    35.1175,
    55.7512,
    48.2100,
    -37.8409,                                      
    49.5833,                                      
    -37.8409,                                      
    29.4393))

cov19.by.lab <- cov19.by.lab %>%
  merge(unique(cov19.by.lab.coords))

cov19.by.lab.pie <- cov19.by.lab %>%
                    reshape2::dcast(city + long + lat + total ~ antigen.epitope, 
                                    value.var = "count", fill = 0)

world <- map_data("world")

set.seed(21)
ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "white", 
    fill = "grey",
    size = 0.2
  ) +
  #geom_point(data = cov19.by.lab, aes(long, lat, size = total)) +
  geom_scatterpie(data = cov19.by.lab.pie, 
                  aes(x=long, y=lat, r = log(total + 1)),
                  colnames(cov19.by.lab.pie)[5:ncol(cov19.by.lab.pie)],
                  color = NA) +
  geom_text_repel(data = cov19.by.lab %>% select(long, lat, city) %>% unique, 
            aes(long, lat, label = city),
            alpha = 0.8) +
  scale_fill_hue(guide="none", "") +
  scale_y_continuous(limits=c(-55, 85)) +
  scale_x_continuous(limits=c(-150, 170)) +
  coord_quickmap() +
  geom_scatterpie_legend(log(cov19.by.lab.pie$total + 1), 
                         x=-125, y=-30, n=3, labeller=function(x) round(exp(x) - 1, 0)) +
  theme_void() +
  theme(legend.position = "none") -> p0

p0

pdf("covid_geo.pdf")
p0
dev.off()
```

---

#### Misc

Number of unique CDR3 sequences per epitope

```{r warning=FALSE, message=FALSE, fig.height=6, fig.width=6}
df.ag = df %>%
  filter(species == "HomoSapiens") %>%
  group_by(gene, mhc.class, antigen.species, antigen.gene, antigen.epitope) %>%
  summarize(cdr3.count = length(unique(cdr3))) %>%
  arrange(-cdr3.count)

ggplot(df.ag, aes(x = cdr3.count, y = rank(-cdr3.count))) +
  #geom_histogram() +
  geom_smooth(method = "lm", 
              color = "red", 
              size = 0.5) +
  geom_point(color = "#253494", alpha = 0.3) +
  scale_x_log10("Unique CDR3s") + scale_y_log10("Epitope rank") +
  facet_grid(mhc.class~gene) +
  theme_classic()
```

Tabular

```{r warning=FALSE, message=FALSE}
colnames(df.ag) = c("TCR chain", "MHC class", "Parent species", "Parent gene", "Epitope sequence", "Unique CDR3")
kable(format = "html", df.ag)
```

Epitopes, parent genes and species - flat table for checking naming

```{r warning=FALSE, message=FALSE}
df %>%
  filter(species == "HomoSapiens") %>%
  select(antigen.species, antigen.gene, antigen.epitope) %>%
  unique %>%
  group_by(antigen.epitope) %>%
  mutate(ambiguous.naming = ifelse(n() > 1, "*", "")) %>%
  arrange(antigen.species, antigen.gene, antigen.epitope) %>%
  kable(format = "html")
```

```{r}
#END
```
