---
title: "VDJdb summary statistics"
author: "Mikhail Shugay"
date: "26-02-2022"
output:
  html_document:
    code_folding: hide
params:
  legacy: "../out/legacy"
  reference_years: "reference_years.tsv"
  annotations: "annotations.tsv"
---

<!--
The release dashboard. Its published fragment -- everything between `!summary_embed_start!` and
`!summary_embed_end!` -- is extracted by `vdjdb.summary.render` and injected into vdjdb-web's
`/overview`, so three properties of the output are load-bearing and asserted by
`summary/check_summary.py`: the base64 images are single unbroken lines, there are no `<div>`
wrappers, and the filename does not change.

The paper figures live in `vdjdb_paper_figures.Rmd`, split off at the marker below. That is what
keeps `maps`, `scatterpie` and `ggrepel` out of this render path.

Every input is a parameter, so the render reads the build under `out/` rather than the stale 2024
copy in `database/`, and it makes no network call:

    Rscript -e 'rmarkdown::render("summary/vdjdb_summary.Rmd")'
    Rscript -e 'rmarkdown::render("summary/vdjdb_summary.Rmd", params = list(legacy = "../out/legacy"))'
-->

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      # Pinned, or the visual fingerprint is noise: the shipped PNGs decode to
                      # 1344x960, 2304x1920, 1152x1920 and 1536x1536, i.e. dpi 96 at retina 2.
                      # Cairo explicitly -- ragg renders fonts differently and would break parity.
                      dev = "png", dev.args = list(type = "cairo"),
                      dpi = 96, fig.retina = 2)
# Every deprecation is reported every time, not once per session, so the upgrade that introduces
# the next one shows it in the render log -- this is what would have surfaced `guide = F`, `size =`
# and `..count..`. NOT "error", which the ROADMAP proposed: measured on ggalluvial 0.12.6 it turns
# that package's own internal `select(axes)` into a hard failure in the first alluvial chunk, so it
# fails on a dependency's debt rather than ours.
options(lifecycle_verbosity = "warning")
# Make sure to keep the default for normal processing.
default_output_hook <- knitr::knit_hooks$get("output")

# Output hooks handle normal R console output.
knitr::knit_hooks$set( output = function(x, options) {

  comment <- knitr::opts_current$get("comment")
  if( is.na(comment) ) comment <- ""
  can_null <- grepl( paste0( comment, "\\s*\\[\\d?\\]" ),
                     x, perl = TRUE)
  do_null <- isTRUE( knitr::opts_current$get("null_prefix") )
  if( can_null && do_null ) {
    # By default R print output aligns at the right brace.
    align_index <- regexpr( "\\]", x )[1] - 1
    # Two cases: start or newline
    re <- paste0( "^.{", align_index, "}\\]")
    rep <- comment
    x <- gsub( re, rep,  x )
    re <- paste0( "\\\n.{", align_index, "}\\]")
    rep <- paste0( "\n", comment )
    x <- gsub( re, rep,  x )
  }

  default_output_hook( x, options )

})

knitr::opts_template$set("kill_prefix"=list(comment=NA, null_prefix=TRUE))
```

```{r message=FALSE, warning=FALSE}
library(knitr)
library(ggplot2)
library(RColorBrewer)
library(data.table)
library(forcats)
library(ggalluvial)
library(circlize)
library(tidyverse)
library(stringr)
library(gridExtra)
library(cowplot)
select = dplyr::select

df = fread(file.path(params$legacy, "vdjdb.slim.txt"), header=T, sep="\t")
```

---

!summary_embed_start!

#### Record statistics by species and TCR chain

General statistics. Note that general statistics are computed using the 'slim' database version. Slim version, for example, lists the same TCR sequence found in several donors/studies only once and selects representative V/J for a given CDR3aa clonotype and the best score across all redundant records.

```{r warning=FALSE, message=FALSE}
# `Studies` counts every reference, not the first of a comma-joined list. Keeping only the first
# field of the split reported 527 of the 636 distinct non-blank references in the slim table, missing
# 109. Rendering one build both ways: human TRA 348 -> 401, human TRB 431 -> 483, mouse TRA 71 -> 84,
# mouse TRB 88 -> 151. `vdjdb.assemble.evidence.support_counts` is the definition this now matches,
# and `tests/unit/test_summary.py` fails if the truncating form comes back.
df.sg = df %>% 
  group_by(species, gene) %>%
  summarize(records = length(complex.id), 
            paired.records = sum(ifelse(complex.id=="0", 0, 1)),
            epitopes = length(unique(antigen.epitope)),
            publications = length(unique(unlist(strsplit(reference.id, ","))))) %>%
  arrange(species, gene)

colnames(df.sg) = c("Species", "Chain", "Records", "Paired records", "Unique epitopes", "Studies")

kable(format = "html", df.sg)
```

#### Record statistics by year

```{r message=FALSE, warning=FALSE, echo=FALSE, cache=FALSE}
# Load data 

dt.vdjdb = fread(file.path(params$legacy, "vdjdb_full.txt"), sep = "\t", fill=T)

dt.vdjdb.s = dt.vdjdb %>%
  filter(species != "MacacaMulatta") %>%
  mutate(tcr_key = paste(v.alpha, j.alpha, cdr3.alpha, v.beta, j.beta, cdr3.beta),
         mhc_key = paste(mhc.a, mhc.b),
         chains = ifelse(cdr3.alpha != "", ifelse(cdr3.beta != "", "paired", "TRA"), "TRB")) %>%
  select(reference.id, tcr_key, mhc_key, chains, antigen.epitope, species) %>%
  unique

# Publication years, read from a committed table rather than fetched.
#
# This block used to call NCBI eutils at render time and cover everything that is not a PubMed id
# with a fourteen-entry literal table last updated in 2021. Measured on the current database that
# table missed 38 of the 52 non-PubMed references -- every PDB structure curated since, and an
# arXiv preprint -- and each one dropped out of the panels below without a warning. The render is
# now offline and deterministic; `uv run vdjdb refs` rebuilds the table by its own pull request.

dt.pubdate = fread(params$reference_years, sep = "\t", header = TRUE) %>%
  transmute(reference.id, pub_date = as.integer(year))

missing.refs = setdiff(unique(dt.vdjdb.s$reference.id[dt.vdjdb.s$reference.id != ""]),
                       dt.pubdate$reference.id)
if (length(missing.refs) > 0) {
  stop(sprintf("%d reference(s) have no year in %s -- run `uv run vdjdb refs`: %s",
               length(missing.refs), params$reference_years,
               paste(head(missing.refs, 5), collapse = ", ")))
}
```

Number of unique TCR sequences, epitopes, MHC alleles and studies by **publication** year, cumulative plots.

```{r message=FALSE, warning=FALSE}
# Cumulative counts by first-appearance year.
#
# The previous form cross-joined every (pub_date, pub_date2) pair -- roughly 7M rows over ~35
# distinct years -- and then ran four `length(unique(x[which(pub_date <= pub_date2)]))` per group.
# That is O(n x years) and the only part of this render that could plausibly exhaust a 16 GB
# runner. A key's first year, tabulated and cumulated, gives the same numbers in one pass.
years = sort(unique(dt.pubdate$pub_date))

cumulative = dt.vdjdb.s %>%
  inner_join(dt.pubdate, by = "reference.id", relationship = "many-to-many") %>%
  select(chains, pub_date, tcr = tcr_key, epi = antigen.epitope,
         ref = `reference.id`, mhc = mhc_key) %>%
  pivot_longer(c(tcr, epi, ref, mhc), names_to = "metric", values_to = "key") %>%
  filter(key != "") %>%
  group_by(chains, metric, key) %>%
  summarise(first_year = min(pub_date), .groups = "drop") %>%
  count(chains, metric, first_year, name = "added") %>%
  complete(chains, metric, first_year = years, fill = list(added = 0)) %>%
  arrange(chains, metric, first_year) %>%
  group_by(chains, metric) %>%
  mutate(total = cumsum(added)) %>%
  ungroup()

# Callouts carry no coordinates (#460). The segment ends at the series value in that year and the
# label floats a fixed fraction of the panel maximum above it, so they stay correct as the database
# grows -- the hardcoded y positions they replace were right for a 2022 database and wrong since.
dt.ann = fread(params$annotations, sep = "\t", header = TRUE) %>%
  mutate(label = gsub("\\\\n", "\n", label))
LABEL_FLOAT = 0.05

callouts = function(metric_name) {
  d = cumulative %>% filter(metric == metric_name)
  a = dt.ann %>% filter(panel == metric_name)
  if (nrow(a) == 0 || nrow(d) == 0) return(NULL)
  top = max(d$total)
  a$value = sapply(a$year, function(y) {
    v = d$total[d$first_year == y]
    if (length(v) == 0) 0 else max(v)
  })
  list(annotate("segment", x = a$year, xend = a$year, y = 0, yend = a$value,
                linetype = "solid", colour = "grey25", linewidth = 0.3),
       annotate("text", x = a$year, y = a$value + LABEL_FLOAT * top, label = a$label,
                hjust = a$hjust, vjust = a$vjust))
}

panel = function(metric_name, title, legend_title = NULL) {
  d = cumulative %>% filter(metric == metric_name)
  p = ggplot(d, aes(x = first_year, y = total, color = chains)) +
    callouts(metric_name) +
    geom_line() +
    geom_point() +
    ylab("") +
    scale_x_continuous("", breaks = seq(1995, max(years), by = 2)) +
    theme_classic() + ggtitle(title) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          axis.line = element_line(linewidth = 0.3))
  if (is.null(legend_title))
    p + scale_color_brewer(palette = "Set1")
  else
    p + scale_color_brewer(legend_title, palette = "Set1") + theme(legend.position = "bottom")
}

p1 = panel("tcr", "Number of unique TCRs")
p2 = panel("epi", "Number of unique epitopes")
p3 = panel("ref", "Number of studies")
p4 = panel("mhc", "Number of MHC alleles", "TCR chain(s)")

# `cowplot::get_legend`, not a grep over grob names. The hand-rolled version looked for a grob
# named exactly "guide-box"; that name still matches on ggplot2 4.0.2, but it is an internal detail
# of the gtable and `tmp$grobs[[integer(0)]]` is the error it throws the release it stops matching.
mylegend = cowplot::get_legend(p4)

grid.arrange(arrangeGrob(p1 + theme(legend.position="none"),
                         p2 + theme(legend.position="none"),
                         p3 + theme(legend.position="none"),
                         p4 + theme(legend.position="none"),
                         nrow=2),
             mylegend, nrow=2,heights=c(10, 1)) #-> PX1

#PX1 %>% plot
#pdf("pubyear.pdf")
#PX1 %>% plot
#dev.off()

#fwrite(dt.vdjdb.s2, "vdjdb_stats_pubyear.txt", sep = "\t")
```

#### Summary by antigen and antigen origin

Representative data for Homo Sapiens

```{r message=FALSE, warning=FALSE}
df.a = df %>% 
  filter(species == "HomoSapiens") %>%
  group_by(antigen.species) %>% #, antigen.gene) %>%
  summarize(records = n(), 
            epitopes = length(unique(antigen.epitope)),
            publications = length(unique(unlist(strsplit(reference.id, ","))))) %>%
  arrange(-records)

colnames(df.a) = c("Parent species", #"Parent gene",
                   "Records", "Unique epitopes", "Studies")
kable(format = "html", df.a)
```

---

#### **COVID-19** data

Summary of antigens and T-cell receptors related to COVID-19 pandemic. Number of records for SARS-CoV-2 epitopes grouped by viral protein and HLA plotted using alluvium plot. Epitopes with less than 30 records in total were not counted.

```{r message=FALSE, warning=FALSE, fig.width=12, fig.height=10}
df %>% 
  filter(species == "HomoSapiens", 
         startsWith(as.character(antigen.species), "SARS-CoV")) %>%
  mutate(mhc.a = str_split_fixed(mhc.a, "[,:]", 2)[,1],
         mhc.b = str_split_fixed(mhc.b, "[,:]", 2)[,1],
         mhc = ifelse(mhc.class == "MHCI", mhc.a, paste0(mhc.a, '/', substr(mhc.b, 7, 15)))) %>%
  group_by(antigen.gene, mhc, antigen.epitope) %>%
  mutate(publications = length(unique(unlist(strsplit(reference.id, ","))))) %>%
  group_by(antigen.gene, mhc, antigen.epitope, gene, publications) %>%
  summarize(records = n()) -> df.c

colnames(df.c) = c("Gene", "HLA", "Epitope", "TCR chain",
                   "Studies",
                   "Records")

ggplot(df.c %>% 
         ungroup %>% arrange(Records) %>% filter(Records >= 30),
       aes(axis1 = Gene,
           axis2 = gsub("HLA-", "", HLA),
           axis3 = substr(Epitope,1,3),
           axis4 = `TCR chain`,
           y = log2(Records))) +
  geom_alluvium(aes(fill = substr(Epitope,1,3) %>% as.factor %>% as.integer), 
                color = "white", alpha = 0.8, curve_type = "sigmoid") +
  geom_stratum(fill = "grey95", color = "white", linewidth =1.0) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_fill_distiller(palette = "Set3", guide="none", "") +
  #scale_fill_hue(guide="none", "") +
  ylab("") + scale_x_discrete(limits = c("Gene", "HLA", "Epitope", "TCR chain"),
                              expand = c(.1, .1),
                              position = "top") +
  theme_void() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x =  element_text(size = 16, color = "black", vjust = -5),
        axis.ticks.x = element_blank(),
        panel.grid.major.y = element_blank())
```

Summary of SARS-CoV-2 epitopes and corresponding TCR alpha and beta chain specificity records (cases with 10+ records)

```{r message=FALSE, warning=FALSE}
kable(format = "html", 
      df.c %>% 
        reshape2::dcast(Gene + HLA + Epitope + Studies ~ `TCR chain`, fill = 0) %>%
        mutate(HLA = gsub("*", ".", HLA, fixed = T)) %>%
        filter(TRA+TRB >= 10) %>%
        arrange(-(TRB+TRA)))
```

---

#### **Self-antigen** data

Summary of T-cell receptors recognizing self-antigens, including antigens linked to utoimmune diseases and potential neoantigen targets for cancer immunotherapy. Number of records for self-antigens grouped by (mutated) human gene and corresponding HLAs are plotted using alluvium plot. Only self-antigens with at least 10 records are shown.

```{r message=FALSE, warning=FALSE, fig.width=12, fig.height=10}
df %>% 
  filter(species == "HomoSapiens", 
         startsWith(as.character(antigen.species), "HomoSapiens")) %>%
  mutate(mhc.a = str_split_fixed(mhc.a, "[,:]", 2)[,1],
         mhc.b = str_split_fixed(mhc.b, "[,:]", 2)[,1],
         mhc = ifelse(mhc.class == "MHCI", mhc.a, paste0(mhc.a, '/', substr(mhc.b, 7, 15)))) %>%
  group_by(antigen.gene, mhc, antigen.epitope) %>%
  mutate(publications = length(unique(unlist(strsplit(reference.id, ","))))) %>%
  group_by(antigen.gene, mhc, antigen.epitope, gene, publications) %>%
  summarize(records = n()) -> df.n

colnames(df.n) = c("Gene", "HLA", "Epitope", "TCR chain",
                   "Studies",
                   "Records")

ggplot(df.n %>% 
         ungroup %>% arrange(Records) %>% filter(Records >= 10),
       aes(axis1 = Gene,
           axis2 = gsub("HLA-", "", HLA),
           axis3 = substr(Epitope,1,3),
           axis4 = `TCR chain`,
           y = log2(Records))) +
  geom_alluvium(aes(fill = substr(Epitope,1,3) %>% as.factor %>% as.integer), 
                color = "white", alpha = 0.8, curve_type = "sigmoid") +
  geom_stratum(fill = "grey95", color = "white", linewidth =1.0) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_fill_distiller(palette = "Accent", guide="none", "") +
  #scale_fill_hue(guide="none", "") +
  ylab("") + scale_x_discrete(limits = c("Gene", "HLA", "Epitope", "TCR chain"),
                              expand = c(.1, .1),
                              position = "top") +
  theme_void() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x =  element_text(size = 16, color = "black", vjust = -5),
        axis.ticks.x = element_blank(),
        panel.grid.major.y = element_blank())
```

Summary of neoantigens and corresponding TCR alpha and beta chain specificity records (cases with 10+ records)

```{r message=FALSE, warning=FALSE}
kable(format = "html", 
      df.n %>% 
        reshape2::dcast(Gene + HLA + Epitope + Studies ~ `TCR chain`, fill = 0) %>%
        mutate(HLA = gsub("*", ".", HLA, fixed = T)) %>%
        filter(TRA+TRB >= 10) %>%
        arrange(-(TRB+TRA)))
```

---

#### Distribution VDJdb confidence scores

Legend: 0 - low confidence, 1 - medium confidence, 2 - high confidence, 3 - very high confidence.

Note that this scoring system is currently deprecated due to large amounts of deep multimer+ repertoire sequencing data and 10X scRNAseq with ImmuDEX dextramer multiplex that are almost impossible to validate independently. We suggest applying methods like TCRNET to [infer high-confidence TCR motifs](https://github.com/antigenomics/vdjdb-motifs) from VDJdb.

```{r message=FALSE, warning=FALSE}
df.score <- df[df$species=='HomoSapiens',] %>%
  group_by(mhc.class, gene, vdjdb.score) %>%
  summarize(total = n())

ggplot(df.score, aes(x=paste(mhc.class, gene, sep = " "), y=total, fill = as.factor(vdjdb.score))) + 
  geom_bar(stat = "identity", position = "dodge", color = "black", linewidth = 0.3) +  
  xlab("") + scale_y_log10("Records") +
  scale_fill_brewer("VDJdb score", palette = "PuBuGn") + 
  theme_classic() +
  theme(legend.position="bottom",
        axis.line = element_line(linewidth = 0.3))
```

---

#### Spectratype

Representative data for Homo Sapiens. CDR3 length distribution (spectratype) is colored by cognate epitope. Second plot shows epitope length distribution for MHC class I and II colored by unique CDR3 (alpha or beta) records.

```{r message=FALSE, warning=FALSE}
df.spe = subset(df, species=="HomoSapiens")

ggplot(df.spe %>% mutate(
  epi_len = nchar(antigen.epitope),
  antigen.epitope = as.factor(antigen.epitope)), 
  aes(x=nchar(as.character(cdr3)))) + 
  geom_histogram(aes(fill = antigen.epitope %>% 
                       fct_reorder(epi_len) %>% as.integer(),
                     group = antigen.epitope %>% 
                       fct_reorder(epi_len)), 
                 bins = 21, size = 0,
                 binwidth = 1, alpha = 0.9, color = NA) + 
  geom_density(aes(y = after_stat(count)), adjust = 3.0, linetype = "dotted") +
  scale_x_continuous(limits = c(5,25), breaks = seq(5,25,5)) + 
  facet_wrap(~gene) + 
  scale_fill_distiller(palette = "Spectral", guide="none", "") +
  #scale_fill_viridis_d(guide="none", direction = -1) +
  xlab("CDR3 length") + ylab("Records") +
  theme_classic() +
  theme(axis.line = element_line(linewidth = 0.3),
        strip.background = element_blank())

# The fill is a Spectral gradient over CDR3-length rank, and the strata are what make the vertical
# gradient inside each bar. Grouping on the CDR3 itself runs stat_bin once per distinct CDR3 --
# 164,131 of them on the human subset -- and stacks that many rectangles per facet, which was 137.82 s
# of a 152 s render. The rank is bucketed into 500 instead: same order, same gradient, and 500 strata
# are already more than a 96-dpi bar can resolve. #648
ggplot(df.spe %>% mutate(
  cdr3_rank = as.factor(cdr3) %>% fct_reorder(nchar(cdr3)) %>% as.integer(),
  cdr3_band = ceiling(cdr3_rank * 500 / max(cdr3_rank))), 
  aes(x=nchar(antigen.epitope))) + 
  geom_histogram(aes(fill = cdr3_band, group = cdr3_band), 
                 size = 0,
                 binwidth = 1, alpha = 0.9, color = NA) + 
  scale_x_continuous(breaks = 7:25) + 
  facet_wrap(.~mhc.class, scales = "free") + 
  scale_fill_distiller(palette = "Spectral", guide="none", "") +
  #scale_fill_viridis_d(guide="none", direction = -1) +
  xlab("Epitope length") + ylab("Records") +
  theme_classic() +
  theme(axis.line = element_line(linewidth = 0.3),
        strip.background = element_blank())
```

#### V(D)J usage and MHC alleles

Representative data for Homo Sapiens, Variable gene. Only Variable genes and MHC alleles with at least 10 records are shown.

```{r message=FALSE, warning=FALSE, fig.height=10, fig.width=6}
df.vhla = df %>% 
  as_tibble %>%
  filter(species == "HomoSapiens" ) %>%
  mutate(id = 1:n()) %>%
  separate_rows(mhc.a, sep = ",") %>%
  separate_rows(mhc.b, sep = ",") %>%
  separate_rows(v.segm, sep = ",") %>%
  mutate(mhc.a.split = str_split_fixed(mhc.a, fixed(":"), n = Inf)[,1],
         mhc.b.split = str_split_fixed(mhc.b, fixed(":"), n = Inf)[,1],
         v.segm.split = str_split_fixed(v.segm, fixed("*"), n = Inf)[,1]) %>%
  group_by(gene, mhc.class, mhc.a.split, mhc.b.split, v.segm.split) %>%
  summarize(records = length(unique(id))) %>%
  group_by(mhc.class, mhc.a.split, mhc.b.split) %>%
  mutate(records.mhc = sum(records)) %>%
  group_by(gene, v.segm.split) %>%
  mutate(records.v = sum(records)) %>%
  filter(records.mhc >= 10, records.v >= 10)
  
#df.vhla$v.segm.split = with(df.vhla, factor(v.segm.split, v.segm.split[order(records.v)]))
#df.vhla$mhc.a.split = with(df.vhla, factor(mhc.a.split, mhc.a.split[order(records.mhc)]))

ggplot(df.vhla, aes(x=gsub("HLA-", "", paste(mhc.a.split, mhc.b.split, sep = " / ")) %>%
                      fct_reorder(records), 
                    y=v.segm.split %>%
                      fct_reorder(records), fill = pmin(records, 1000))) +
  geom_tile() +
  scale_fill_gradientn("Records", colors=colorRampPalette(brewer.pal(9, 'PuBuGn'))(32), 
                       trans="log", breaks = c(1, 10, 100, 1000)) +
  xlab("") + ylab("") +
  facet_grid(gene~mhc.class, scales="free", space="free") +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        axis.text.y = element_text(size = 8),
        panel.grid.major = element_blank(),
        legend.position = "right",
        axis.line = element_line(linewidth = 0.3),
        strip.background = element_blank())
```

Circos plot for correspondence between human TRBV genes and MHC class I alleles for links supported by more than 50 records in terms of the ratio of observed to expected records. Band width and color (from red:highest to light yellow:lowest) are scaled proportional to co-occurrence matrix value divided by row (TRBV) and column (HLA) sums.

```{r message=FALSE, warning=FALSE, fig.width=8, fig.height=8}
#https://jokergoo.github.io/circlize_book
df.vhla.1b <- df.vhla %>%
  filter(records >= 50) %>%
  filter(gene == "TRB", mhc.class == "MHCI") %>%
  mutate(mhc.a.split = substr(mhc.a.split, 5, nchar(mhc.a.split)),
         v.segm.split = paste0("Vb", substr(v.segm.split, 5, nchar(v.segm.split)))) %>%
  reshape2::dcast(mhc.a.split ~ v.segm.split, value.var = "records", fill = 0)
rownames(df.vhla.1b) <- df.vhla.1b$mhc.a.split
df.vhla.1b$mhc.a.split <- NULL
df.vhla.1b <- df.vhla.1b %>% as.matrix()
df.vhla.1b <- (df.vhla.1b %>% 
  sweep(1, rowSums(df.vhla.1b), `/`) %>%
  sweep(2, colSums(df.vhla.1b), `/`)) * sum(df.vhla.1b)

grid.col <- setNames(c(colorRampPalette(brewer.pal(9, 'YlGn'))(nrow(df.vhla.1b)) %>% rev, 
                       colorRampPalette(brewer.pal(9, 'PuBu'))(ncol(df.vhla.1b))), 
                     union(rownames(df.vhla.1b), colnames(df.vhla.1b)))
band.col <- colorRamp2(range(df.vhla.1b), 
                       colorRampPalette(brewer.pal(9, 'YlOrRd'))(2), 
                       transparency = 0.2)

chordDiagram(df.vhla.1b, annotationTrack = "grid", preAllocateTracks = 1, 
             col = band.col,
             grid.col = grid.col,
             grid.border = "black",
             big.gap = 20)

circos.trackPlotRegion(track.index = 1, panel.fun = function(x, y) {
  xlim = get.cell.meta.data("xlim")
  ylim = get.cell.meta.data("ylim")
  sector.name = get.cell.meta.data("sector.index")
  circos.text(mean(xlim), ylim[1] + .1, sector.name, 
              facing = "clockwise", niceFacing = T, adj = c(0, 0.5), 
              cex = 0.65)
}, bg.border = NA)

title("TRBV ~ HLA-I")

circos.clear()
```

---

#### Detailed summary for HLA

Representative data for Homo Sapiens MHC class I and II

```{r message=FALSE, warning=FALSE}
df.m = df %>% 
  as_tibble %>%
  filter(species == "HomoSapiens") %>%
  mutate(id = 1:n()) %>%
  separate_rows(mhc.a, sep = ",") %>%
  separate_rows(mhc.b, sep = ",") %>%
  mutate(mhc.a.split = str_split_fixed(mhc.a, fixed(":"), n = Inf)[,1],
         mhc.b.split = str_split_fixed(mhc.b, fixed(":"), n = Inf)[,1]) %>%
  group_by(mhc.a.split, mhc.b.split) %>%
  summarize(records = length(unique(id)), 
            antigens = length(unique(antigen.epitope)),
            publications = length(unique(unlist(strsplit(reference.id, ","))))) %>%
  arrange(-records)
  
colnames(df.m) = c("First chain", "Second chain", "Records", "Unique epitopes", "Studies")
kable(format = "html", df.m)
```

!summary_embed_end!
